Title Goes Here

### 2023-02-07 inst/paper/_discussion.qmd
add co-author comments round 2

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids and salicylic
acid from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry (UHPLC-MS/MS), a method that was validated by co-occurrence
of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We extend findings from previous studies on alkaloids in archaeological samples
by extracting multiple different alkaloids from dental calculus, including
cannabinoids, cocaine, nicotine, cotinine, caffeine, theophylline, and salicylic
acid in multiple individuals. Of these compounds, the cannabinoids and cocaine
could not be replicated in a second run on different samples from the same individuals.
We discuss the implications of these findings in light of historical and
archaeological evidence for the consumption of these drugs.

@@

<!-- correlations with skeletal indicators -->
We explored the relationship between detected compounds and various skeletal
indicators, such as pathological and dental lesions, preservation, and pipe notches.
We found some evidence to suggest that preservation of the skeleton influences
the recovery of compounds from the dental calculus, with well-preserved skeletons
potentially serving as a better target for sampling.  
We found no correlation between the number of pipe notches
and the concentration of nicotine and cotinine, suggesting that our ability
to detect tobacco consumption in dental calculus does not necessarily rely on
targeting frequent smokers; here, we consider individuals with multiple pipe notches
as likely to have been heavy smokers.  
We found a positive correlation between CMS and nicotine, which may be
indicative of the impact tobacco smoking had on the respiratory health of the
Beemster inhabitants.
Tobacco smoke may play a significant role in diseases
of the upper respiratory tract, including CMS [@rehImpactTobacco2012]. Although
the mechanisms by which smoking increases the risk of infections is not fully
understood, solid evidence has been presented linking tobacco smoke to
increased mucosal permeability and impairment of mucociliary clearance
[@arcaviCigaretteSmoking2004]. Such changes, together with an altered immunologic
response, are thought to predispose to the development of CMS
[@slavinDiagnosisManagement2005].  
We also observed a moderate positive correlation between CMS and caffeine which
contradicts previous research linking chronic coffee consumption with a positive
effect on the respiratory system, suggesting a preventive association between
caffeine intake and pneumonia
[e.g. @alfaroChronicCoffee2018; @kondoAssociationCoffee2021].
However, while the
lower respiratory tract seems to benefit from chronic coffee consumption, it is
possible that elevated caffeine intake impacts mucosal moisture due to its
dehydrating effect [@maughanCaffeineIngestion2003], thereby exposing individuals
to greater risk of sinus infection.

@@

<!-- limitations of the platform show which questions would be unanswered -->
Since we targeted individuals with moderate-to-large calculus deposits,
it is likely a biased sample. The presence of calculus may increase risk of premature death
[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
be associated with dental calculus build-up) is a risk-factor for
respiratory diseases, if periodontal and respiratory pathogens
enter the blood stream
[@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
In our sample, the percentage of CMS
(`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
is lower than in another (more representative) male sample (44.1%)
[@casnaUrbanizationRespiratory2021], and the caries percentage is similarly lower
in our sample (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`) than
a more representative sample (22.9%) [@lemmersMiddenbeemster2013].  
We used the presence/absence of a pipe notch and concurrent detection of tobacco
as a crude estimate of the accuracy of the method, which we found to be around
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
This is a very rough estimate, as the presence of a pipe notch is likely not a
perfect indicator of whether or not someone consumed tobacco.
Dental calculus is also
more transient than for example bone, as it can be mechanically removed,
intentionally or unintentionally, during life, eliminating
all trace of the alkaloids consumed prior to its removal.  
Quantitation of the detected compounds may have limited value in archaeological
samples due to degradation, and will greatly affect our correlations related to
concentration.
Following burial, compound stability over
time will play a large role, as will microbial degradation of compounds by bacteria
and funghi in soil [@liuNicotinedegradingMicroorganisms2015], as well as
the soil environment, such as temperature, pH, and oxygen availability
[@mackiePreservationMetaproteome2017; @lindholstLongTerm2010].  
The detected quantity of a compound will also depend on the quantity in dental
calculus during life, which is largely controlled by quantity of consumption,
how often the
calculus was disrupted/removed, metabolic breakdown of the compound, and inter-
and intra-individual factors related to stages of biofilm formation, maturation,
and mineralisation
[@lustmannScanningElectron1976; @velskoMicrobialDifferences2019; @zijngeBiofilmArchitecture2010].
In short, this means it is not really possible to detect the absence of a compound.
The absence of a compound is not evidence of absence of consumption. This complicates
the interpretation of our results.
We have attempted to minimise errors occurring due to this limitation by including
a relatively large sample of individuals. Although given the relatively low
detection rate seen in tobacco, this remains a major limitation, and will likely
be compounded by increasing antiquity of the samples.
The replication of samples increases the evidence of absence if the compound
was absent in both batches.

Future studies should explore how sampling from various types of teeth
and their position in the mouth affects the probability of a compound becoming
entrapped in dental calculus.
This may also be related to properties within the oral cavity, as well as chemical
properties of the compounds, which facilitate or reduce the incorporation-potential,
and which incorporation pathways are more likely for a given compound.  
We only targeted drugs that were included in the forensic toxicological
screenings, and therefore only covered a limited number of the potential compounds
that could be of interest for exploring past diets and medicinal treatments. The
list of targeted compounds can be expanded as we discover more potential targets
based on which specific compounds/metabolites are more likely to be incorporated
and preserved in dental calculus.  
We found a strong correlation between nicotine and cotinine, but
a lesser correlation between caffeine and theophylline. This is likely because
caffeine and theophylline have multiple plausible sources in our population, but
more research is needed on the relationship between parent compounds and their
metabolites in dental calculus.  
There is an increasing interest in using oral fluid as a means of detecting alkaloids
in living
individuals due to the non-invasive nature of the testing compared to blood
and urine sampling [@coneSalivaTesting1993; @valenDetermination212017].
These *in vivo* studies are a valuable
source of method validation and can help determine the feasibility of detecting
certain alkaloids in oral fluid and, subsequently, dental calculus. Archaeologists,
though, will likely be responsible for exploring dental calculus specific
incorporation and retention of alkaloids, as well as their long-term preservation
in the burial environment.<!-- move to discussion in disseration: Here, *in vitro* studies
may provide valuable insights-->

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method is in detecting food and other items
that are consumed in a manner that does not allow macroscopic detection. Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, with others including chewing, drinking, cigars,
and snuff [@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
so methods like the one presented here
are suitable for exploring tobacco consumption in an entire society, rather
than a trivial subset of past populations.<!--admittedly, this study did just that...-->

>>>>

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids and salicylic
acid from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry (UHPLC-MS/MS), a method that was validated by co-occurrence
of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We extend findings from previous studies on alkaloids in archaeological samples
by extracting multiple different alkaloids from dental calculus, including
nicotine, cotinine, caffeine, theophylline, and salicylic
acid in multiple individuals. The detection of these compounds was
solidified in a replication analysis on different samples from the same
individuals.
Cocaine and multiple cannabinoids were also detected during the first
analysis, but were not replicated.
We discuss the implications of these findings in light of historical and
archaeological evidence for the consumption of these drugs.

@@

<!-- correlations with skeletal indicators -->
We explored the relationship between detected compounds and various skeletal
indicators, such as pathological and dental lesions, preservation, and pipe notches.
We found some evidence to suggest that preservation of the skeleton influences
the recovery of compounds from the dental calculus, with well-preserved skeletons
potentially serving as a better target for sampling.  
We found a positive correlation between CMS and nicotine, which may be
indicative of the impact tobacco smoking had on the respiratory health of the
Beemster inhabitants.
Tobacco smoke may play a significant role in diseases
of the upper respiratory tract, including chronic maxillary sinusities
[@rehImpactTobacco2012]. Although
the mechanisms by which smoking increases the risk of infections is not fully
understood, solid evidence has been presented linking tobacco smoke to
increased mucosal permeability and impairment of mucociliary clearance
[@arcaviCigaretteSmoking2004]. Such changes, together with an altered immunologic
response, are thought to predispose to the development of chronic maxillary sinusitis
[@slavinDiagnosisManagement2005].  
We also observed a moderate positive correlation between chronic maxillary sinusitis and caffeine which
contradicts previous research linking chronic coffee consumption with a positive
effect on the respiratory system, suggesting a preventive association between
caffeine intake and pneumonia
[e.g. @alfaroChronicCoffee2018; @kondoAssociationCoffee2021].
However, while the
lower respiratory tract seems to benefit from chronic coffee consumption, it is
possible that elevated caffeine intake impacts mucosal moisture due to its
dehydrating effect [@maughanCaffeineIngestion2003], thereby exposing individuals
to greater risk of sinus infection.

@@

<!-- limitations of the platform show which questions would be unanswered -->
Because we targeted individuals with moderate-to-large calculus deposits,
it is likely a biased sample. The presence of calculus may increase the
risk of premature death
[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
be associated with dental calculus build-up) is a risk-factor for
respiratory diseases, if periodontal and respiratory pathogens
enter the bloodstream
[@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
In our sample, the percentage of chronic maxillary sinusitis
(`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
is lower than in another (more representative) male sample (44.1%)
[@casnaUrbanizationRespiratory2021], and the caries percentage is similarly lower
in our sample (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`) than
a more representative sample (22.9%) [@lemmersMiddenbeemster2013].  
We used the presence/absence of a pipe notch and concurrent detection of tobacco
as a crude estimate of the accuracy of the method, which we found to be around
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
This is a very rough estimate, as the presence of a pipe notch is likely not a
perfect indicator of whether or not someone consumed tobacco.
Dental calculus is also
more transient than for example bone, as it can be mechanically removed,
intentionally or unintentionally, during life, eliminating
all trace of the alkaloids consumed prior to its removal.  
Quantitation of the detected compounds may have limited value in archaeological
samples due to degradation, and will greatly affect our correlations related to
concentration.
Following burial, compound stability over
time will play a large role, as will microbial degradation of compounds by bacteria
and fungi in soil [@liuNicotinedegradingMicroorganisms2015], as well as
the soil environment, such as temperature, pH, and oxygen availability
[@mackiePreservationMetaproteome2017; @lindholstLongTerm2010].  
The detected quantity of a compound will also depend on the quantity in dental
calculus during life, which is largely controlled by quantity of consumption,
how often the
calculus was disrupted/removed, metabolic breakdown of the compound, and inter-
and intra-individual factors related to stages of biofilm formation, maturation,
and mineralisation
[@lustmannScanningElectron1976; @velskoMicrobialDifferences2019; @zijngeBiofilmArchitecture2010].
In short, this means it is not really possible to detect the absence of a compound.
The absence of a compound is not evidence of absence of consumption. This complicates
the interpretation of our results.
We have attempted to minimise errors occurring due to this limitation by including
a relatively large sample of individuals and replicating our analysis. Although
given the relatively low detection rate seen in tobacco, this remains a major
limitation, and will likely be compounded by increasing antiquity of the samples.

Future studies should explore how sampling from various types of teeth
and their position in the mouth affects the probability of a compound becoming
entrapped in dental calculus.
This may also be related to properties within the oral cavity, as well as chemical
properties of the compounds, which facilitate or reduce the incorporation-potential,
and which incorporation pathways are more likely for a given compound.  
We only targeted drugs that were included in the forensic toxicological
screenings, and therefore only covered a limited number of the potential compounds
that could be of interest for exploring past diets and medicinal treatments. The
list of targeted compounds can be expanded as we discover more potential targets
based on which specific compounds/metabolites are more likely to be incorporated
and preserved in dental calculus.  
There is an increasing interest in using oral fluid as a means of detecting alkaloids
in living
individuals due to the non-invasive nature of the testing compared to blood
and urine sampling [@coneSalivaTesting1993; @valenDetermination212017].
These *in vivo* studies are a valuable
source of method validation and can help determine the feasibility of detecting
certain alkaloids in oral fluid and, subsequently, dental calculus. Archaeologists,
though, will likely be responsible for exploring dental calculus specific
incorporation and retention of alkaloids, as well as their long-term preservation
in the burial environment.<!-- move to discussion in disseration: Here, *in vitro* studies may provide valuable insights-->

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->

<!-- rephrase -->
While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method lies
in the ability to detect dietary and other compounds that were incorporated
via multiple
consumption pathways that are not detected by other methods.
Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, with others including chewing, drinking, cigars,
and snuff [@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
so methods like the one presented here
are suitable for exploring tobacco consumption in an entire society, rather
than a trivial subset of past populations.<!--admittedly, our sample was just that...-->
Combined with other methods, it can also give us a more complete picture
of dietary patterns and medicinal/recreational plant-use in the past by capturing
multiple possible incorporation pathways of dietary (and other) compounds.

### 2023-02-07 inst/paper/_intro.qmd
add co-author comments round 2

<!-- define alkaloids -->
One such biomarker is alkaloids, a plant-derived group of compounds. Many alkaloids
have important medicinal and psychoactive effects in humans, and their direct
detection, or detection of their metabolites, is of great interest to archaeologists.
Previous studies have successfully recovered alkaloids in archaeological
contexts, including ceramics [@smithDetectionOpium2018], pipes
[@raffertyCurrentResearch2012], human hair
[@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013], and even
dental calculus employing both targeted [@eerkensDentalCalculus2018] and
untargeted approaches
[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].
Especially nicotine, the principal alkaloid in tobacco leaves,
has been widely studied in the archaeological record due to its apparent stability
and ability to survive over long periods of time
[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].

<!--how are compounds incorporated in dental calculus-->
Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion as alkaloids and other compounds are transferred from
plasma to saliva, and then into the oral cavity through the salivary glands
in the hours-to-days following consumption [@coneInterpretationOral2007]. The
relation to plasma
is why there is often a close correlation between presence (not concentration)
of drugs in oral fluid and blood
[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
The second pathway allows the
detection of parent compounds and their metabolites that not only are consumed orally,
but also any other mode
of consumption or administration, as long as it is excreted through saliva<!--cite e.g. injection or topical The second pathway may also be more
detectable, as the excretion of compounds through saliva may take hours to days
whereas compounds spend less time in the oral cavity during direct consumption-->.

@@

<!--brief description of method validation-->
In this study we use a ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-MS/MS) method that was developed in a previous study on dental
calculus from cadavers and validated by comparing the results to compounds detected
in the blood of the same individuals.
All compounds that were detected in the blood were also detected in dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long period of time [@sorensenDrugsCalculus2021].
<!--summary of results-->
We were able to detect both parent compounds and metabolites, including caffeine,
nicotine, theophylline, and cotinine, in the dental calculus of individuals from
a 19th century Dutch population from Middenbeemster. By detecting these compounds
we are able to show the consumption
of tea and coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches in the
dentition.

>>>>

<!-- define alkaloids -->
One such collection of biomarkers is alkaloids, a plant-derived group of compounds.
Many alkaloids
have important medicinal and psychoactive effects in humans, and their direct
detection, or detection of their metabolites, is of great interest to archaeologists.
Previous studies have successfully recovered alkaloids in archaeological
contexts, including ceramics [@smithDetectionOpium2018], pipes
[@raffertyCurrentResearch2012], human hair
[@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013], and even
dental calculus employing both targeted [@eerkensDentalCalculus2018] and
untargeted approaches
[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].
Especially nicotine, the principal alkaloid in tobacco leaves,
has been widely studied in the archaeological record due to its apparent stability
and ability to survive over long periods of time
[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].

<!--how are compounds incorporated in dental calculus-->
Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion as alkaloids and other compounds are transferred from
plasma to saliva, and then into the oral cavity through the salivary glands
in the hours to days following consumption [@coneInterpretationOral2007]. The
relation to plasma
is why there is often a close correlation between presence (not concentration)
of drugs in oral fluid and blood
[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
The second pathway allows the identification of parent compounds that are not
consumed orally, as long as they, or their metabolites, are excreted through saliva.
<!--cite e.g. injection or topical The second pathway may also be more
detectable, as the excretion of compounds through saliva may take hours to days
whereas compounds spend less time in the oral cavity during direct consumption-->.

@@

<!--brief description of method validation-->
In this study we use a ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-MS/MS) method that was developed in a previous study on dental
calculus from cadavers and validated by comparing the results to compounds detected
in the blood of the same individuals [@sorensenDrugsCalculus2021].
All compounds that were detected in the blood were also detected in dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long period of time [@sorensenDrugsCalculus2021].
<!--summary of results-->
We were able to detect both parent compounds and metabolites, including caffeine,
nicotine, theophylline, and cotinine, in the dental calculus of individuals from
a 19th century Dutch population from Middenbeemster. By detecting these compounds
we are able to show the consumption
of tea and coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches in the
dentition.

### 2023-02-07 inst/paper/_materials.qmd
add co-author comments round 2

<!-- Demographics and description of MB11 sample, selection criteria. -->
The sample consists of `r nrow(demography)` individuals from Middenbeemster,
a 19th century rural Dutch site.
Of the `r nrow(demography)` individuals that were analysed,
`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
replicated using leftover calculus from the original DNA study.
<!--in the same lab? different lab? -->
The village of Middenbeemster and the surrounding Beemsterpolder was established
in the beginning of the 17th century, when the Beemster lake was drained to
create more farm land, mainly for the cultivation of cole seeds (de Vries 1978).
In 1615, a decision was made to build a church, and
construction started in 1618 (Hakvoort 2013). The excavated cemetery is associated
with the Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
[@lemmersMiddenbeemster2013]. Archival documents are available for those buried
between AD 1829 and 1866, when the majority of individuals were interred
[@palmerActivityReconstruction2016]. The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the industrial
revolution [@aten400JaarBeemster2012; @palmerActivityReconstruction2016].

To reduce the number of factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49 years). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable females, and
`r nrow(filter(demography, sex == "f"))` female [@fig-sample-demography].
We selected males due to a higher
occurrence of dental calculus deposits than females (unpublished observation).

>>>>

<!-- Demographics and description of MB11 sample, selection criteria. -->
The sample consists of `r nrow(demography)` individuals from Middenbeemster,
a 19th century rural Dutch site.
The village of Middenbeemster and the surrounding Beemsterpolder was established
in the beginning of the 17th century, when the Beemster lake was drained to
create more farmland, mainly for the cultivation of cole seeds (de Vries 1978).
In 1615, a decision was made to build a church, and
construction started in 1618 (Hakvoort 2013). The excavated cemetery is associated
with the Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
[@lemmersMiddenbeemster2013]. Archival documents are available for those buried
between AD 1829 and 1866, when the majority of individuals were interred
[@palmerActivityReconstruction2016]. The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the industrial
revolution [@aten400JaarBeemster2012; @palmerActivityReconstruction2016].

To reduce the number of potentially confounding factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49 years). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable females, and
`r nrow(filter(demography, sex == "f"))` female (@fig-sample-demography).
We selected males due to a higher
occurrence of pipe notches and dental calculus deposits than females (unpublished observation).

### 2023-02-07 inst/paper/_methods.qmd
add co-author comments round 2

New dental calculus samples were taken under sterile conditions in a positive
pressure laminar flow hood in a dedicated dental calculus lab at Leiden University.
The surface of the tooth was lightly brushed with a sterile, disposable toothbrush
to get rid of surface contaminants. A sterile dental curette was then used to
scrape calculus from the tooth onto weighing paper, which was transferred to 1.5 ml
Eppendorf tubes.
Where possible, previously sampled material was used to prevent repeated sampling
of individuals. The sampling methods can be found in the published material
[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
The samples were sent to the Department of Forensic Medicine at Aarhus University
for ultra-high-performance liquid chromatography-tandem mass spectrometry (UHPLC-MS/MS)
analysis.

@@

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985], and divided into the following categories:
early young adult (EYA, 18-24 years), late young adult (LYA, 25-34 years), middle adult (MA, 
35-49 years), old adult (Old, 50+ years).

Pathological conditions and lesions that occur frequently in the population were
included in the analysis. Data were
dichotomised to presence/absence to allow statistical analysis.
Osteoarthritis (OA) was considered present in cases where eburnation was visible
on one or more joint surfaces.
Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
formation on the margin of the superior and inferior surfaces of the vertebral
body.
Cribra orbitalia (CO) was diagnosed based on presence of pitting on the superior
surface of the orbit. No distinction was made between active or healing lesions.
Degenerative disc disease (DDD), or spondylosis, is identified as a large diffuse
depression of the
superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
Schmorl's nodes (SN) are identified as any cortical depressions on the surface of
the vertebral body.
Data on chronic maxillary sinusitis (CMS) from @casnaUrbanizationRespiratory2021 were
included in this study to assess the relationship between upper respiratory
diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
Lesions associated with CMS as defined by
@boocockMaxillarySinusitis1995 were
recorded for each individual and classified as "pitting", "spicule-type bone
formation", "remodeled spicules", or "white pitted bone". CMS was scored as absent
when the sinus presented smooth surfaces with little or no associated pitting.

Caries ratios were calculated by dividing the number of lesions by the number
of teeth scored, resulting in a single caries ratio per individual. If the surface
where the lesion originated is not visible, i.e. if the lesion covered multiple
surfaces, this was scored as "crown".
Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
on the lingual, buccal (labial), and interproximal surfaces of each tooth. Only
one score was used for the combined interproximal surfaces, resulting in three
scores per tooth (when surfaces are intact), and four calculus indices per individual;
upper anterior (UA), upper posterior (UP), lower anterior (LA), lower posterior (LP).
Each index was calculated by dividing
the sum of calculus scores for each surface by the total number of surfaces scored
in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
was not included [@greeneQuantifyingCalculus2005].
Periodontitis was scored on a visual four-stage (0-3) scoring system according to
distance from CEJ of each tooth to alveolar bone [@maatManualPhysical2005].

>>>>

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985],
and divided into the following categories:
early young adult (18-24 years), late young adult (25-34 years), middle adult (
35-49 years), old adult (50+ years).

Pathological conditions and lesions that occur frequently in the population were
included in the analysis. Data were
dichotomised to presence/absence to allow statistical analysis.
Osteoarthritis was considered present in cases where eburnation was visible
on one or more joint surfaces.
Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
formation on the margin of the superior and inferior surfaces of the vertebral
body.
Cribra orbitalia was diagnosed based on the presence of pitting on the superior
surface of the orbit. No distinction was made between active or healing lesions.
Degenerative disc disease, or spondylosis, is identified as a large diffuse
depression of the
superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
Schmorl's nodes are identified as any cortical depressions on the surface of
the vertebral body.
Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
included in this study to assess the relationship between upper respiratory
diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
Lesions associated with chronic maxillary sinusitis as defined by
@boocockMaxillarySinusitis1995 were
recorded for each individual and classified as "pitting", "spicule-type bone
formation", "remodeled spicules", or "white pitted bone".
chronic maxillary sinusitis was scored as absent
when the sinus presented smooth surfaces with little or no associated pitting.

Caries ratios were calculated by dividing the number of lesions by the number
of teeth scored, resulting in a single caries ratio per individual. If the surface
where the lesion originated is not visible, i.e. if the lesion covered multiple
surfaces, this was scored as "crown".
Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
on the lingual, buccal (and labial), and interproximal surfaces of each tooth. Only
one score was used for the combined interproximal surfaces, resulting in three
scores per tooth (when surfaces are intact), and four calculus indices per individual;
upper anterior, upper posterior, lower anterior, lower posterior.
Each index was calculated by dividing
the sum of calculus scores for each surface by the total number of surfaces scored
in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
was not included [@greeneQuantifyingCalculus2005].
Periodontitis was scored on a visual four-stage (0-3) scoring system according to
distance from cemento-enamel junction of each tooth to alveolar bone [@maatManualPhysical2005].

### Calculus sampling

Where possible, we used material that had already been sampled for a previous
study to prevent unnecessary repeated sampling of individuals. 
Calculus from the previous study was
sampled in a dedicated ancient DNA laboratory at the Laboratories of Molecular
Anthropology and Microbiome Research in Norman, Oklahoma, U.S.A,
using established ancient DNA protocols. More details on the methods can be found
in the published articles
[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
Of the `r nrow(demography)` individuals that were originally included in our sample,
`r nrow(filter(metadata, replicated == TRUE))`
were replicated in a separate analysis only using calculus from the previous study.  
New dental calculus samples were taken under sterile conditions in a positive
pressure laminar flow hood in a dedicated dental calculus lab at Leiden University.
The surface of the tooth was lightly brushed with a sterile, disposable toothbrush
to get rid of surface contaminants. A sterile dental curette was then used to
scrape calculus from the tooth onto weighing paper, which was transferred to 1.5 ml
Eppendorf tubes.
All calculus samples were sent to the Department of Forensic Medicine at Aarhus University
for ultra-high-performance liquid chromatography-tandem mass spectrometry (UHPLC-MS/MS)
analysis.

### UHPLC-MS/MS

@@
